---
title: "replicationGames"
author: "Steph, Sam, Iggy"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load Packages
library(readr) # read csv file in
library(dplyr) # wrangling
library(tidyr) # Remove NAs in merged data
library(lubridate) # dates
library(sf) # foor reading .shp data

```

```{r}
# Setup - Load in data file
# set working directory as source file location
# Read in upzipped .csv data from "input" folder within

df1 <- read_csv("input/Data_2000-2016_county_monthly_combined.csv")

df2 <- rbind(
    read_csv("input/CompMort_01_08.csv"),
    read_csv("input/CompMort_09_16.csv")
)

# Merge data
df3 <- merge(
    mutate(df2, GEOID = `County Code`, year = Year),
    df1,
    all.x = TRUE,
    by = c("GEOID", "year")
) 

# Clean data
df4 <- df3 %>%
    mutate(
        CVD.adj = as.numeric(`Age Adjusted Rate`, na.rm = TRUE),
        GEOID = as.factor(GEOID)
        ) %>% # change column classes
    dplyr::select(
        c(
            GEOID,
            year_month,
            year,
            month,
            PM25,
            NO2,
            O3,
            Tmean,
            dewT,
            pop.total,
            pop.Male,
            pop.Female,
            pop.White,
            pop.Black,
            pop.Hispanic,
            CVD.adj
        )
    ) %>% # Select required columns
    drop_na() # Remove missing values


US.county <- st_read("input/US_continental_counties_3103/US_continental_counties_3103.shp")


```

## Iggy Section

code goes here

```{r cars}
summary(cars)
```

## Steph Section

### Figure 1ab

#### Figure 1ab (Their Code)

```{r}
# using packages dplyr, lubridate, & sf

# SJH added
# re assign df1 "Data_2000-2016_county_monthly_combined.csv" to data
data <- df1


## average PM2.5 (Figure 1a)
PM25.ave <- data %>%
  group_by(GEOID) %>%
  summarise(PM25_mean = mean(PM25))

## PM2.5 change (Figure 1b)
PM25.2000 <- data %>%
  filter(year==2000) %>%
  group_by(GEOID) %>%
  summarise(PM25_2000 = mean(PM25))

PM25.2016 <- data %>%
  filter(year==2016) %>%
  group_by(GEOID) %>%
  summarise(PM25_2016 = mean(PM25))

PM25.change <- PM25.2000 %>%
  left_join(PM25.2016, by=c("GEOID")) %>%
  mutate(PM25_chg = (PM25_2016 - PM25_2000)/PM25_2000 * 100) %>%
  dplyr::select(GEOID, PM25_chg)


## join with shapefile
PM25.ave$GEOID <- as.character(PM25.ave$GEOID)
PM25.change$GEOID <- as.character(PM25.change$GEOID)

US.county.F1 <- US.county %>%
  left_join(PM25.ave, by="GEOID") %>%
  left_join(PM25.change, by="GEOID")

```

#### Figre 1ab (SJH code)

-   Note - "ArcGIS Pro was used to make the final maps." (Github \> Analysis Description.pdf \> 2.2 Descriptive Maps)

    -   Attempting to use the sf package to map information instead.

    -   <https://r-spatial.github.io/sf/articles/sf5.html>

    -   <https://r-spatial.github.io/sf/articles/sf5.html#ggplot2>

```{r}
# Using sf package to plot geography/Ca
# Basic plot of the USA
plot(st_geometry(US.county.F1))

# basic ggplot test of PM25_mean plot
ggplot() + 
  geom_sf(data = US.county.F1, aes(fill = PM25_mean)) + 
  #scale_y_continuous(breaks = 34:36)
  theme_bw()
```

##### Data prep for Fig 1a & 1b

```{r}
# using 
US.county.F1ab <- US.county.F1 %>%
  mutate(
    PM25_mean_rank = case_when(
      PM25_mean > 12 ~ "12.01-13.74",
      PM25_mean > 9 ~ "9.00-12.00",
      PM25_mean > 6 ~ "6.00-9.00",
      PM25_mean > 3 ~ "3.00-6.00",
      PM25_mean <= 3 ~ "2.15-3.00",
      TRUE ~ "F" # Default case: if none of the above are true
    )
  ) %>%
    mutate(
    PM25_chg_rank = case_when(
      PM25_chg > 20 ~ "20.00-52.99",
      PM25_chg > 0 ~ "0.00-20.00",
      PM25_chg > -20 ~ "-20.00-0.00",
      PM25_chg > -40 ~ "-40.01--20.00",
      PM25_chg <= -40 ~ "-60.43--40.00",
      TRUE ~ "F" # Default case: if none of the above are true
    )
  )

# Colors used in figure 1a
fig_1a_colors <- c("#006837", "#86CB67", "#FFFFBF", "#F98E52", "#A50026")

# Colors used in figure 1b
fig_1b_colors <- c("#313695", "#74ADD1", "#E0F3F8", "#F98E52", "#A50026")
```

##### Figure 1a replication 

```{r}

fig_1a <- ggplot() + 
  geom_sf(data = US.county.F1ab, aes(fill = PM25_mean_rank)) +
  scale_fill_manual(name = "Average PM2.5 concentration",
                    breaks=c("2.15-3.00", "3.00-6.00", "6.00-9.00",
                             "0.00-20.00", "12.01-13.74"),
                    values = c("#006837", "#86CB67", "#FFFFBF",
                               "#F98E52", "#A50026")) + 
  labs(title = "Figure 1a") +
  theme_bw()

fig_1a
```

##### Figure 1b Replication

```{r}

fig_1b <- ggplot() + 
  geom_sf(data = US.county.F1ab, aes(fill = PM25_chg_rank)) +
  scale_fill_manual(name = "Change in PM2.5 concentration (%)",
                    breaks= c("-60.43--40.00", "-40.01--20.00", "-20.00-0.00",
                             "0.00-20.00", "20.00-52.99"),
                    values = c("#313695", "#74ADD1", "#E0F3F8",
                      "#F98E52", "#A50026")) + 
  labs(title = "Figure 1b") +
  theme_bw()

fig_1b
```

### Figure 1c

#### Figure 1c (Their Code)

```{r}
library(dplyr)
library(ggplot2)
library(cowplot)
library(MetBrewer)

## filter for cardiovascular mortality data
mortality.CVD  <- mortality %>%
  filter(ICD.category == "I" & ICD.number >= 0 & ICD.number <= 99.99) %>%
  dplyr::select(year_month, res_FIPS, race, hispanic) %>%
  rename(GEOID = res_FIPS) %>%
  mutate(race_cat = case_when(race==1 & hispanic==0 ~ "non-Hispanic White",
                              race==2 & hispanic==0 ~ "non-Hispanic Black",
                              hispanic==1 ~ "Hispanic"),
         race_cat = factor(race_cat, levels = c("non-Hispanic White","non-Hispanic Black","Hispanic"))) %>%
  dplyr::select(GEOID, year_month, race_cat) %>%
  filter(is.na(race_cat)==F)

rm(mortality)


## select PM2.5 data
data.AP <- data %>%
  dplyr::select(GEOID, year_month, year, PM25)

## 12-month moving average of PM2.5 concentration
data.list <- split(data.AP, as.factor(data$GEOID))
for (j in 1:length(data.list)){
  data.list[[j]][,paste0("PM25_lag0",11)]<- tsModel::runMean(data.list[[j]]$PM25, c(0:11))
}

data.AP.01to16 <- bind_rows(data.list) %>%
  filter(year >= 2001) %>%
  dplyr::select(-PM25,-year)


## individual-level long-term PM2.5 exposure
AP.individual <- mortality.CVD %>%
  left_join(data.AP, by=c("GEOID","year_month")) %>%
  left_join(data.AP.01to16, by=c("GEOID","year_month")) 

## make the violin plot
theme_set(theme_cowplot())

plot.individual.12M <- ggplot(AP.individual, aes(x=race_cat, y=PM25_lag011, fill=race_cat)) +
  geom_violin() +
  scale_fill_met_d("Kandinsky", direction = 1) +
  geom_boxplot(width=0.1, color="grey", alpha=0.2)+
  xlab("")+
  ylab("12-month moving average of PM2.5") +
  theme(legend.position="none")
```

## Sam Section

Code goes here
